---
title: "Plastic ingestion meta-analysis: data preparation and checking"
author: Andrew MacDonald with updates from Natalie Cooper
format: html
editor: source
bibliography: references.bib
---

*** before final run of models need to update tree and check which species need to be removed from data because
they are not in the tree, and which species have NAs for diet
why do some have n total = 0???***

This document performs data cleaning and checking for this study. Note that I (NC) strip down the phylogeny from the full bird phylogeny in a different script (data-preparation/03-get-phylogenetic-trees-BBtree)

### Reading data

```{r}
# Load libraries
library(ape) # for phylogeny import
library(tidyverse) # for data import and manipulation
library(here) # to ensure paths work
library(naniar) # to look for NAs
library(geiger) # to check tips and species match in tree and data
library(janitor) # clean dataset
```

#### Import tree

This is the new Claramont et al 2025 bird tree using the Clements 2024 taxonomy.

```{r}
# Read in tree
seabird_tree <- read.nexus(here::here("data/seabird-tree-Claramont2025_2025-09-11.nex"))

# Look at it
str(seabird_tree) # should have 385 tips
plot(seabird_tree, no.margin = TRUE, cex = 0.5, type = "fan")
```

#### Read in data

These are datasets from Sept 2025. ***Update to most recent***.
```{r}
# Read in trait data and plastic data
data_traits <- read_csv(here::here("data-raw/trait-data-2025-09-24.csv"))
data_plastic <- read_csv(here::here("data-raw/plastic-ingestion-data-2025-09-12.csv"))

# Look at the data
glimpse(data_traits) # 391 species of seabird
glimpse(data_plastic) 
```

## Data transformations and calculations

We do some data transformations before we get to work:

-   median year of a study is recalculated, to give a decimal number
-   some columns are dropped
-   body size is standardized with TWO standard deviations, as suggested by @gelman2008
-   trait and plastic datasets are merged
-   add a column called tips which is the Binomial name with "_" to match the tree

### Calculate median date

```{r median-calc}
## recalculate median so we have a number or NA
data_plastic <- data_plastic %>%  
  rowwise() %>% 
  mutate(Date_median = median(c(Date_FirstYR, Date_LastYR))) %>% 
  ungroup()
```

### Merge datasets

```{r data-merge}
# Join the two data files
data_all <- data_plastic %>% 
  left_join(data_traits, by = "Clements2024_Binomial")
```

### Convert factors to binary
Make the "Yes/No" variables into 0 or 1 (helps to clarify how they work in the models). 0 = No, 1 = Yes

```{r binary-factors}
## Factors that are "YES" vs "No" can just be binary.-------------
# Set factors
data_all <- data_all %>% 
  mutate(across(c(Carrion, Cephalopods, Crustaceans, Fish,
                  Other.Inverts, Anthropogenic.Subsidy,
                  Pelagic.Specialist,
                  AS.Fishery, AS.Refuse, AS.Agriculture,
                  aerial_pursuit, bottom_feeding, diving, 
                  plunging, scavenging, skimming, surface_seizing, terrestrial_feeding),
                ~ if_else(.x == "Yes", true = 1, false = 0)))
```

### Add tips column to help with matching to the tree

```{r}
data_all <-
  data_all %>%
  mutate(tips = gsub(" ", "_", data_all$Clements2024_Binomial))
```

### Body mass standardisation
Body size standardized with TWO standard deviations, as suggested by @gelman2008
```{r}
data_all <-
  data_all %>%
  mutate(body_mass_2sd = (Body.Mass.g - mean(Body.Mass.g, na.rm = TRUE)) / (2 * sd(Body.Mass.g, na.rm = TRUE)))
```

### Freeflying

Rather than recognizing the fine distinctions between different life-history stages, we are aggregating into just two life history stages: free-flying and not (i.e. nestbound). This is important because freeflying birds are foraging for themselves, which nestbound birds are consuming food foraged by parents. However, this is in fact a study-level variable, as studies typically focus on either one the other life history stage.

```{r}
data_all <- data_all %>% 
  mutate(freeflying = case_when(
    Age.Class == "Nest-bound" ~ 0L,
    Age.Class %in% c("Subadult/Adult", "Adult", "Subadult") ~ 1L,
    Age.Class == "Mixed" ~ NA_integer_
  ))
```

# Simplify sampling method by choosing primary method where two are listed

```{r}
data_all <- data_all %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Bolus necropsy", "Bolus")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Bolus/Necropsy", "Bolus")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Bolus/lavage", "Bolus")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Lavage/Necropsy", "Lavage")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Regurgitation/Necropsy", "Regurgitation")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Regurgitation/Lavage", "Regurgitation")) %>%
  mutate(Sampling_method = str_replace(Sampling_method, "Scat/Necropsy", "Scat"))
```

### Select columns for the models

```{r}
# n  = sample size
# F0 = frequency of occurrence
data_full <- data_all %>%
  dplyr::select(Clements2024_Binomial,
    tips, Order, 
    # main response variable and sample sizes
    n, n_plastic, FO, Plastic,
    ## species level stuff
    Body.Mass.g, body_mass_2sd, 
    # diet as a factor
    Diet, 
    ## diet breakdown
    Carrion, Cephalopods, Crustaceans, Fish,
    Other.Inverts, Anthropogenic.Subsidy,
    Pelagic.Specialist,
    AS.Fishery, AS.Refuse, AS.Agriculture,
    aerial_pursuit, bottom_feeding,
    diving, plunging, scavenging, skimming, surface_seizing, terrestrial_feeding,
    ## Study-level predictors
    Source, Date_median, Feeding.method,
    MarineRegion, freeflying, UniqueID_Plastic.study, Sampling_method)

```

## Clean up the names using janitor

```{r}
data_full <- data_full %>% clean_names()
```


### Drop species that are not in the phylogeny

This should just be seven records where only the genus was recorded. Specifically: Catharacta_spp. Diomedea_spp., Fregetta_spp., Gavia_sp.,Macronectes_spp., Pachyptila_spp., Phaethon_sp. For code used to prune the Claramont 2025 tree see `data-preparation/03-get-phylogenetic-trees-BBtree.R`

```{r}
check <- name.check(seabird_tree, data_full, data.names = data_full$tips)
# List mismatches
unique(check$data_not_tree)

# Remove the species not in the tree from the data
data_full <-
  data_full[is.na(match(data_full$tips, check$data_not_tree)),]
```

## Aggregating the data

Some studies separate birds into small groups (i.e. within-study sampling locations). However, for the purposes of meta-analyses, we are treating all observations as exchangeable (i.e. as replicates) if they:

-   are collected by the same authors (Source)
-   in the same Marine Region
-   in the same Year

Note that we have to include all other variables so they aren't lost in the aggregation.

```{r}
# Group by all variables except unique study ID
data_agg <- data_full |> 
  group_by(clements2024_binomial, tips, order,
           # study level traits
           freeflying,
           source, 
           date_median, 
           marine_region, 
           sampling_method,
           # species level traits
           body_mass_g,
           body_mass_2sd, 
           main_diet = diet,
           carrion, 
           fish,
           cephalopods,
           crustaceans,
           other_inverts,
           as_fishery,
           as_agriculture,
           as_refuse,
           anthropogenic_subsidy,
           pelagic_specialist,
           feeding_method,
           aerial_pursuit, 
           bottom_feeding,
           diving, 
           plunging, 
           scavenging, 
           skimming, 
           surface_seizing, 
           terrestrial_feeding) %>%
# Summarise the n_plastic etc for each
summarise(n_plastic = sum(n_plastic, na.rm = TRUE),
          n_total = sum(n, na.rm = TRUE),
          FO = n_plastic/n_total, .groups = "drop")
```

### Fix errors in FO

- In some cases n_total and n_plastic are both 0, so FO is NaN. Make this zero.
- In one case after aggregation n_plastic is higher than n_total meaning FO is Inf, remove this entry.

```{r}
data_agg <- 
  data_agg %>%
  # where n_total and n_plastic are 0, FO should be 0
  mutate(FO = case_when(n_total == 0 & n_plastic == 0 ~ 0,
                       TRUE ~ as.numeric(FO))) %>%
  # where there is n_plastic but no n_total, FO is Inf - remove this entry
  filter(!is.infinite(FO))
```

### A few quick checks

```{r}
# Check for missing data
# This produces a plot showing missing data
naniar::vis_miss(data_agg)
```

## Write out full list of species with plastic studies of some kind

Get species totals for numbers of studies, how many with plastic etc.
```{r}
species_list <- 
  data_agg %>%
  group_by(tips) %>%
  summarise(total_plastic = sum(n_plastic), 
            total_n = sum(n_total),
            species_FO = total_plastic / total_n,
            n_studies = n()) %>%
  mutate(plastic = case_when(species_FO > 0 ~ "yes", species_FO == 0 ~ "no"))

# Add back in the species without any studies

no_studies <- data.frame(tips = check$tree_not_data, 
                         total_plastic = rep(NA, length(check$tree_not_data)),
                         total_n = rep(NA, length(check$tree_not_data)),
                         species_FO = rep(NA, length(check$tree_not_data)),
                         n_studies = rep(0, length(check$tree_not_data)),
                         plastic = rep(NA, length(check$tree_not_data)))

# combine
full_species_list <- rbind(species_list, no_studies)

write_csv(full_species_list, file = here::here("data/full_species_list_2025-09-16.csv"))
```

### Without NAs

Remove a small number of annoying NAs which we aren't able to work with. We lose a few studies, which include few species each, and one species with no diet info. 

```{r}
## Remove entries with missing years and diet and sampling method info
# Also remove the nestbound individuals
data_agg_noNAs <- data_agg %>%
  filter(!is.na(date_median)) %>%
  filter(!is.na(main_diet)) %>%
  filter(!is.na(carrion)) %>%
  filter(!is.na(body_mass_g)) %>%
  filter(!is.na(sampling_method)) %>%
  filter(freeflying == 1 | is.na(freeflying))
```

## Write out full dataset

```{r}
write_csv(data_agg_noNAs, file = here::here("data/plastic-data-aggregated_2025-09-15.csv"))
```

## Write out a phylogeny which only includes species in the dataset for the analyses

```{r}
# Check for missing taxa
check <- name.check(seabird_tree, data_agg_noNAs, data.names = data_agg_noNAs$tips)

check$data_not_tree # should be zero
check$tree_not_data # species with no plastic data

# Drop tips
seabird_tree_plastic <- drop.tip(seabird_tree, check$tree_not_data)
str(seabird_tree_plastic) # Should have 230 tips
length(unique(data_agg_noNAs$tips)) # Should be 230

# Write to file
write.nexus(seabird_tree_plastic, file = here::here("data/seabird_tree_plastic_2025-09-15.nex"))
```