---
title: "Plastic ingestion meta-analysis: data preparation and checking"
author: Andrew MacDonald with updates from Natalie Cooper
format: html
editor: source
bibliography: references.bib
---

*** before final run of models need to update tree and check which species need to be removed from data because
they are not in the tree, and which species have NAs for diet***

This document performs data cleaning and checking for this study. Note that I (NC) strip down the phylogeny from the full bird phylogeny in a different script (data-preparation/04-get-phylogenetic-trees-mcc.R)

### Reading data

```{r}
# Load libraries
library(ape) # for phylogeny import
library(tidyverse) # for data import and manipulation
library(here) # to ensure paths work
library(naniar) # to look for NAs
library(geiger) # to check tips and species match in tree and data
```

Import tree

This is the Hackett backbone version of the Jetz tree for now, with taxonomy updates to current HBW v 8. Update to clements tree when a dated version is available.

```{r}
# Read in tree
tree.seabird <- read.nexus(here::here("data/seabird-tree-HBW-2024-10-11.nex"))

# Look at it
str(tree.seabird) # should have 380 tips
plot(tree.seabird, no.margin = TRUE, cex = 0.5, type = "fan")
```

Read in data

These are datasets from Oct 2024. Update to most recent.
```{r}
# Read in trait data and plastic data
data0.1 <- read_csv(here::here("data-raw/trait-data-2024-10-02.csv"))
data0.2 <- read_csv(here::here("data-raw/plastic-ingestion-data-2024-10-02.csv"))

# Look at the data
glimpse(data0.1) # 388 species of seabird
glimpse(data0.2) # ? plastic studies. Note that one Unique plastic ID was removed when we excluded scaup.
```

## Data transformations and calculations

We do some data transformations before we get to work:

-   median year of a study is recalculated, to give a decimal number

-   some columns are dropped

-   body size is standardized with TWO standard deviations, as suggested by @gelman2008

-   data0.1 and data0.2 are merged

-   add a column called tips which is the HBW binomial with _ to match the tree

### Calculate median date

```{r median-calc}
## recalculate median so we have a number or NA
data0.2 <- data0.2 |> 
  rowwise() |> 
  mutate(Date_median = median(c(Date_FirstYR, Date_LastYR))) |> 
  ungroup()
```

### Merge datasets

```{r data-merge}
# Join the two data files
data2_joined <- data0.2 |> 
  left_join(data0.1, by = "HBWv8.1_Binomial")
```

### Convert factors to binary
Make the "Yes/No" variables into 0 or 1 (helps to clarify how they work in the models). 0 = No, 1 = Yes

```{r binary-factors}
## Factors that are "YES" vs "No" can just be binary.-------------
# Set factors
data2_binary <- data2_joined |> 
  mutate(across(c(Carrion.NonFishVertebrates, Cephalopods, Crustaceans, Fish,
                  Other.Inverts, Anthropogenic.Subsidy,
                  pelagic_specialist_CR,DMS.responsive_Dell.Ariccia,
                  AS.Fishery, AS.Refuse, AS.Agriculture),
                ~ if_else(.x == "Yes", true = 1, false = 0)))
```

## Add tips column

```{r}
data2_tips <-
  data2_binary %>%
  mutate(tips = gsub(" ", "_", data2_binary$HBWv8.1_Binomial))
```

## Select columns for the models

```{r}
# n  = sample size
# F0 = frequency of occurrence
data2 <- data2_tips |> 
  select(HBWv8.1_Binomial,
    HBWv8.1_Order, tips,
    # main response variable and sample sizes
    n, n_plastic, FO, Plastic,
    ## species level stuff
    Body.Mass.g,
    # diet as a factor
    Diet, 
    ## diet breakdown
    Carrion.NonFishVertebrates, Cephalopods, Crustaceans, Fish,
    Other.Inverts, Anthropogenic.Subsidy,
    pelagic_specialist_CR,
    AS.Fishery, AS.Refuse, AS.Agriculture,
    # olfactory trap
    DMS.responsive_Dell.Ariccia,
    ## Study-level predictors
    Source,
    Date_median,
    Feeding.Method.Simple,
    MarineRegion,
    # Site,
    Age.Class, UniqueID_Plastic.study
    ) |>
  # body size standardized with TWO standard deviations, as suggested by @gelman2008
  mutate(body_mass_2sd = (
           Body.Mass.g - mean(Body.Mass.g, na.rm = TRUE)) / (2 * sd(Body.Mass.g, na.rm = TRUE))
         )
```

### Freeflying

Rather than recognizing the fine distinctions between different life-history stages, we are aggregating into just two life history stages: free-flying and not (ie nestbound). This is important because freeflying birds are foraging for themselves, which nestbound birds are consuming food foraged by parents. However, this is in fact a study-level variable, as studies typically focus on either one the other life history stage.

```{r}
data2_freefly <- data2 |> 
  mutate(is_freeflying = case_when(
    Age.Class == "Nest-bound" ~ 0L,
    Age.Class %in% c("Subadult/Adult", "Adult", "Subadult") ~ 1L,
    Age.Class == "Mixed" ~ NA_integer_
  ))

```

### Drop species that are not in the phylogeny

Check this with the new phylogeny!

```{r}
check <- name.check(tree.seabird, data2_freefly, data.names = data2_freefly$tips)
unique(check$data_not_tree)

data2_matched <-
  data2_freefly %>% 
  filter(HBWv8.1_Binomial != "Aphrodroma brevirostris")
```


### Exclude study with no plastic info

Check this with the new data!

```{r}
#data2_matched$UniqueID_Plastic.study[which(is.na(data2_matched$Plastic))]

data2_complete <- data2_matched |> 
  filter(UniqueID_Plastic.study != 551)
```

## Aggregating within marine basins

Some studies separate birds into small groups (i.e. within-study sampling locations).
However, for the purposes of meta-analyses, we are treating all observations as exchangeable (i.e. as replicates) if they:

-   are collected by the same authors
-   in the same Marine Region
-   in the same year

This aggregates the dataset considerably.

```{r}
data2_aggregated <- data2_complete |> 
  group_by(HBWv8.1_Binomial, tips,
           # study level traits
           is_freeflying,
           Source, 
           Date_median, 
           MarineRegion, 
           # species level traits
           body_mass_2sd, 
           Carrion.NonFishVertebrates, 
           Fish,
           Cephalopods,
           Crustaceans,
           Other.Inverts,
           AS.Refuse,
           pelagic_specialist_CR,
           Feeding.Method.Simple,
           DMS.responsive_Dell.Ariccia
           ) |> 
  summarize(n_plastic = sum(n_plastic, na.rm = TRUE),
            n_total = sum(n, na.rm = TRUE),
            n_na_plastic = sum(is.na(n_plastic)), .groups = "drop")
```

### A few quick checks

```{r}
# Check for missing data
# This produces a plot showing missing data
# Note that some plastic studies have FO but not n or n_plastics, 
# and a few have Plastic (Yes/No) but no FO, n or n_plastic
# If they have none of these they should be excluded.
naniar::vis_miss(data2_aggregated)
```

### Remove missing species from the phylogeny

Drop species from the tree that are not in the dataset
```{r}
# Check for missing taxa
check <- name.check(tree.seabird, data2_aggregated, data.names = data2_aggregated$tips)

check$data_not_tree # should be 0
length(check$tree_not_data) # all the seabirds we don't have plastic records for 

# Drop tips
tree_plastic <- drop.tip(tree.seabird, check$tree_not_data)
str(tree_plastic)
```

## Write out data and tree

```{r}
readr::write_csv(data2_aggregated, 
                 file = here::here("data/plastic-data-aggregated.csv"))

write.nexus(tree_plastic, file = here::here("data/plastic-tree.nex"))
```

### Without NAs

Also write a version without a small number of annoying NAs which we aren't able to work with:

```{r}
## Remove entries with missing years and missing diet data
data2_noNAs <- data2_aggregated |> 
  filter(!is.na(Date_median)) |> 
  filter(!is.na(Fish))

readr::write_csv(data2_noNAs, file = here::here("data/plastic-data-aggregated-noNA.csv"))

## what is lost in this filtering?
data2_aggregated |> 
  anti_join(data2_noNAs, by = join_by(HBWv8.1_Binomial, Source)) |> 
  count(Source)
```

We lose a few studies, which include few species each. We also drop one bird species for which very little diet information is available. 

Also make a tree without these species with NAs...
```{r}
# Check for missing taxa
check <- name.check(tree_plastic, data2_noNAs, data.names = data2_noNAs$tips)

check$data_not_tree # should be zero
check$tree_not_data # species with NAs that have been excluded

# Drop tips
tree_plastic_noNA <- drop.tip(tree_plastic, check$tree_not_data)
str(tree_plastic_noNA)

# Write to file
write.nexus(tree_plastic_noNA, file = here::here("data/plastic-tree-noNA.nex"))
```

## Add back in species that are seabirds but we don't have plastic data for

This is so we can make phylogeny plots showing all the species in the tree, even ones
without existing data

*** Currently this doesn't work as I want it to. ***

```{r, eval=FALSE}
fulldata <- 
  data2_aggregated %>%
  # full join to the trait data
  full_join(data0.1, by = "HBWv8.1_Binomial") %>%
  # ensure tips column is complete and matches tree with _ not " "
  mutate(tips = gsub(" ", "_", .$HBWv8.1_Binomial))
  
  
  

# Check for missing taxa in tree/data
check <- name.check(tree.seabird, fulldata, data.names = fulldata$tips)
check$data_not_tree # should be zero
check$tree_not_data # species with NAs that have been excluded

# Drop tips
tree_plastic_full <- drop.tip(tree.seabird, check$tree_not_data)
str(tree_plastic_full) # should have 380 tips

# Write to file
write.nexus(tree_plastic_full, file = here::here("data/plastic-tree-full.nex"))

# What species need to be removed from data that are not in the tree?
check$data_not_tree

# Remove missing species from dataset
fulldata_match <-
  fulldata %>%
  filter(tips != "Aphrodroma_brevirostris" &
        tips != "Bulweria_bifax"  &        
        tips != "Fregetta_lineata" &        
        tips != "Oceanites_pincoyae" &    
        tips != "Pinguinus_impennis" &    
        tips != "Pterodroma_rupinarum" &
        tips != "Puffinus_bryani" & 
        tips != "Urile_perspicillatus")

# Write to file
readr::write_csv(fulldata_match, file = here::here("data/plastic-data-full.csv"))
```
