---
title: "Exploring variation in the data"
format: html
editor: visual
---

***not many species are well studied and still fewer are susceptible - check this in the previous ms***

# Explore the data that have been collected

```{r}
# Load libraries
library(tidyverse) # for plots and data wrangling
library(patchwork) # for panel plots
```

Read in the data and combine diet categories
```{r, message = FALSE}
data_noNA <- readr::read_csv(file = here::here("data/plastic-data-aggregated-noNA.csv"))

# Read in tree
tree <- read.nexus(here::here("data/plastic-tree-noNA.nex"))

# Unite the diet categories
data_noNA_trait <- data_noNA |> 
  unite(col = "diet_type", Carrion.NonFishVertebrates:AS.Refuse)
```

## Some summary stats

```{r}
# How many species?
length(unique(data_noNA$HBWv8.1_Binomial))
```

```{r}
# How many sources?
length(unique(data_noNA$Source)) 
```

For each species, we calculated the mean FO, weighted by the sample size (i.e., weighted mean FO). 

```{r}
susceptible <-
data_noNA %>%
  mutate(FO = (n_plastic/n_total) * 100) %>%
  group_by(HBWv8.1_Binomial) %>%
   summarise(FO_weighted_mean = weighted.mean(FO))

ggplot(susceptible, aes(x = FO_weighted_mean)) +
  geom_density() +
  xlab("FO weighted mean") +
  theme_bw()
```

1. How many species are "well-studied"?

We classify ‘well-studied’ species as those with ≥ 100 sampled birds across ≥ 3 records (i.e., multiple studies, years, locations and/or age classes).
```{r}
well_studied <- 
data_noNA %>%
  group_by(HBWv8.1_Binomial) %>%
  summarise(records = n(), sampled = sum(n_total)) %>%
  filter(records >=3 & sampled >= 100)

well_studied %>% summarise(n())
```

2. How many well studied species are susceptible to plastic ingestion?

We classify well-studied species as susceptible to plastic ingestion if weighted mean FO ≥ 50%.

```{r}
susceptible_well_studied <-
data_noNA %>%
  mutate(FO = (n_plastic/n_total) * 100) %>%
  group_by(HBWv8.1_Binomial) %>%
   summarise(FO_weighted_mean = weighted.mean(FO, n_total),
             records = n(), sampled = sum(n_total)) %>%
  filter(records >=3 & sampled >= 100 & FO_weighted_mean >= 50)

susceptible_well_studied %>% summarise(n())
```


## Make distributional plots of each variable

```{r}
# Frequency of occurrence
fo <- 
  ggplot(data_noNA, aes(x = n_plastic/n_total)) +
  geom_density() +
  xlab("Frequency of occurrence of plastic") +
  theme_bw()

# Body mass
bm <-
  ggplot(data_noNA, aes(x = body_mass_2sd)) +
  geom_density() +
  xlab("Scaled body mass") +
  theme_bw()

# Free flying or not
ff <- 
  ggplot(data_noNA_trait, aes(x = as.factor(is_freeflying), fill = as.factor(is_freeflying))) +
  geom_bar() +
  xlab("") +
  scale_x_discrete(labels = c("Nest-bound", 
                              "Free flying", "Mixed")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A", "#508991")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) 

# Median date
dd <- 
  ggplot(data_noNA_trait, aes(x = Date_median)) +
  geom_density() +
  xlab("Median date") +
  theme_bw()

# Marine region
mr <- 
ggplot(data_noNA_trait, aes(x = as.factor(MarineRegion), fill = as.factor(MarineRegion))) +
  geom_bar() +
  xlab("Marine region") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#E5E1EE","#DFFDFF","#90BEDE","#68EDC6","#90F3FF","#3D3B8E")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))  
 
# Fish
f <-
ggplot(data_noNA, aes(x = as.factor(Fish), fill = as.factor(Fish))) +
  geom_bar() +
  xlab("Fish") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))  
 
 # Carrion.NonFishVertebrates
c <-
ggplot(data_noNA, aes(x = as.factor(Carrion.NonFishVertebrates), fill = as.factor(Carrion.NonFishVertebrates))) +
  geom_bar() +
  xlab("Carrion (non fish)") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A")) 
 
# Cephalopods
cp <-
  ggplot(data_noNA, aes(x = as.factor(Cephalopods), 
                       fill = as.factor(Cephalopods))) +
  geom_bar() +
  xlab("Cephalopods") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
# Crustaceans
cr <-
ggplot(data_noNA, aes(x = as.factor(Crustaceans), fill = as.factor(Crustaceans))) +
  geom_bar() +
  xlab("Crustaceans") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
 # Other inverts
oi<- 
ggplot(data_noNA, aes(x = as.factor(Other.Inverts), fill = as.factor(Other.Inverts))) +
  geom_bar() +
  xlab("Other invertebrates") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
 # AS.Refuse
ar <-
ggplot(data_noNA, aes(x = as.factor(AS.Refuse), fill = as.factor(AS.Refuse))) +
  geom_bar() +
  xlab("Refuse") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

# Feeding method simple
fm <-
ggplot(data_noNA, aes(x = as.factor(Feeding.Method.Simple), fill = as.factor(Feeding.Method.Simple))) +
  geom_bar() +
  xlab("Feeding method") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","#43AA8B",
                               "#577590")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) 
 
# pelagic_specialist_CR
ps <- 
ggplot(data_noNA, aes(x = as.factor(pelagic_specialist_CR), fill = as.factor(pelagic_specialist_CR))) +
  geom_bar() +
  xlab("Pelagic specialist") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

# DMS.responsive
dms <-
ggplot(data_noNA, aes(x = as.factor(DMS.responsive_Dell.Ariccia), fill = as.factor(DMS.responsive_Dell.Ariccia))) +
  geom_bar() +
  xlab("DMS responsive") +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
```

Fit patchwork grids for plots

Species-level variables, i.e. feeding, body size etc.
```{r}
(f + cp + cr) / (c + oi + ar) / (dms + ps) / (fm + bm)
ggsave(file = here("output/species-level-variables.png"), height = 8)
```

Study level variables
```{r}
 (fo + dd) / (ff + mr)
ggsave(file = here("output/study-level-variables.png"), height = 6)
```

```{r}
ggplot(data_noNA, aes(x = as.factor(is_freeflying), y = n_plastic/n_total)) +
  geom_boxplot()
```

