---
title: "Exploring variation in the data"
format: html
editor: visual
---

***not many species are well studied and still fewer are susceptible - check this in the previous ms***

# Explore the data that have been collected

## Load libraries

```{r}
# Load libraries
library(tidyverse) # for plots and data wrangling
library(patchwork) # for panel plots
library(here)
```

## Read in the data and tree

```{r, message = FALSE}
# Read in the data
seabird_data <- read_csv(file = here::here("data/plastic-data-aggregated_2025-09-15.csv"))
# Read in tree
tree <- read.nexus(here::here("data/seabird_tree_plastic_2025-09-15.nex"))
```

## Unite the diet categories

```{r}
# Unite the diet categories
seabird_data <- seabird_data %>%
  unite(col = "diet_type", carrion:as_refuse, remove = FALSE)
```

## Some summary stats

```{r}
# How many species?
length(unique(seabird_data$clements2024_binomial))
```

```{r}
# How many sources?
length(unique(seabird_data$source)) 
```

### How many species are "well-studied"?

We classify ‘well-studied’ species as those with ≥ 100 sampled birds across ≥ 3 records (i.e., multiple studies, years, locations and/or age classes).

```{r}
well_studied <- 
seabird_data %>%
  group_by(clements2024_binomial) %>%
  summarise(records = n(), sampled = sum(n_total)) %>%
  filter(records >=3 & sampled >= 100)

well_studied %>% summarise(n())
```

## Make distributional plots of each variable

```{r}
# Body mass
bm <-
  ggplot(seabird_data, aes(x = log(body_mass_g))) +
  geom_density() +
  xlab("ln(Body mass) (g)") +
  theme_bw()

# main diet
diet <-
ggplot(seabird_data, aes(x = as.factor(main_diet), fill = as.factor(main_diet))) +
  geom_bar() +
  xlab("Main diet") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#F94144","#F3722C",
                               "#F9C74F","#90BE6D",
                               "#577590")) +
  scale_x_discrete(labels = c("carrion", "cephalopods", "crustaceans", "fish", "other inverts")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) 

# Fish
f <-
ggplot(seabird_data, aes(x = as.factor(fish), fill = as.factor(fish))) +
  geom_bar() +
  xlab("Fish") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))  
 
 # Carrion.NonFishVertebrates
c <-
ggplot(seabird_data, aes(x = as.factor(carrion), fill = as.factor(carrion))) +
  geom_bar() +
  xlab("Carrion (non fish)") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A")) 
 
# Cephalopods
cp <-
  ggplot(seabird_data, aes(x = as.factor(cephalopods), 
                       fill = as.factor(cephalopods))) +
  geom_bar() +
  xlab("Cephalopods") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
# Crustaceans
cr <-
ggplot(seabird_data, aes(x = as.factor(crustaceans), fill = as.factor(crustaceans))) +
  geom_bar() +
  xlab("Crustaceans") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
 # Other inverts
oi<- 
ggplot(seabird_data, aes(x = as.factor(other_inverts), fill = as.factor(other_inverts))) +
  geom_bar() +
  xlab("Other invertebrates") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))
 
 # AS.Refuse
ar <-
ggplot(seabird_data, aes(x = as.factor(as_refuse), fill = as.factor(as_refuse))) +
  geom_bar() +
  xlab("Refuse") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

af <-
ggplot(seabird_data, aes(x = as.factor(as_fishery), fill = as.factor(as_fishery))) +
  geom_bar() +
  xlab("Fisheries") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

aa <-
ggplot(seabird_data, aes(x = as.factor(as_agriculture), fill = as.factor(as_agriculture))) +
  geom_bar() +
  xlab("Agriculture") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

# Feeding method
fm <-
ggplot(seabird_data, aes(x = as.factor(feeding_method), fill = as.factor(feeding_method))) +
  geom_bar() +
  xlab("Feeding method") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","#43AA8B",
                               "#577590")) +
  scale_x_discrete(labels = c("aerial pursuit", "bottom feeding", "diving", 
                              "plunging", "scavenging", "skimming", "surface seizing")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) +
  ylim(0,800)
 
# Pelagic Specialist
ps <- 
ggplot(seabird_data, aes(x = as.factor(pelagic_specialist), fill = as.factor(pelagic_specialist))) +
  geom_bar() +
  xlab("Pelagic specialist") +
  ylim(0, 1650) +
  scale_x_discrete(labels = c("No", "Yes")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#F3722C","#577590"))

# DMS.responsive
#dms <-
#ggplot(data_noNA, aes(x = as.factor(DMS.responsive_Dell.Ariccia), fill = #as.factor(DMS.responsive_Dell.Ariccia))) +
  #geom_bar() +
 # xlab("DMS responsive") +
  #scale_x_discrete(labels = c("No", "Yes")) +
  #theme_bw() +
  #theme(legend.position = "none") +
  #scale_fill_manual(values = c("#74B3CE","#09BC8A"))
```

Fit patchwork grids for plots

Species-level variables, i.e. feeding, body size etc.

```{r}
(fm + bm + diet) / (f + c + cp) / (cr + oi + aa) / (af + ar + ps)
#ggsave(file = here("output/species-level-variables.png"), height = 8)
```

Study level variables

```{r}
# Frequency of occurrence
fo <- 
  ggplot(seabird_data, aes(x = FO)) +
  geom_histogram(fill = "lightblue", bins = 10) +
  xlab("Frequency of occurrence of plastic") +
  theme_bw()

# Free flying or not
ff <- 
  ggplot(seabird_data, aes(x = as.factor(freeflying), fill = as.factor(freeflying))) +
  geom_bar() +
  xlab("Age class") +
  scale_x_discrete(labels = c("Free flying", "Mixed")) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74B3CE","#09BC8A"))

# Median date
dd <- 
  ggplot(seabird_data, aes(x = date_median)) +
  geom_histogram(fill = "sienna1", bins = 100) +
  xlab("Date of study (median)") +
  theme_bw()

# Marine region
mr <- 
ggplot(seabird_data, aes(x = as.factor(marine_region), fill = as.factor(marine_region))) +
  geom_bar() +
  xlab("Marine region") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","#43AA8B",
                               "#577590")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))  

# Sampling method
sm <- 
ggplot(seabird_data, aes(x = as.factor(sampling_method), fill = as.factor(sampling_method))) +
  geom_bar() +
  xlab("Sampling method") +
  theme_bw() +
  theme(legend.position = "none") +
    scale_fill_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","grey")) +
   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))  
```

```{r}
 (dd + ff) / (mr + sm)
ggsave(file = here("output/study-level-variables.png"), height = 6)
```

Some simple correlative plots

```{r}
# Get summary stats for each grouping variable
sumdatSM <-
  seabird_data %>%
  group_by(sampling_method) %>%
  summarise(meanFO = mean(FO), seFO = sqrt(var(FO)/n()))

sumdatDate <-
  seabird_data %>%
  group_by(date_median) %>%
  summarise(meanFO = mean(FO), seFO = sqrt(var(FO)/n()))

sumdatDiet <-
  seabird_data %>%
  group_by(main_diet) %>%
  summarise(meanFO = mean(FO), seFO = sqrt(var(FO)/n()))

sumdatFM <-
  seabird_data %>%
  group_by(feeding_method) %>%
  summarise(meanFO = mean(FO, na.rm = TRUE), seFO = sqrt(var(FO, na.rm = TRUE)/n()))
```

Study level variables
```{r}
# Plots
# Sampling method
ggplot(seabird_data, aes(x = as.factor(sampling_method), y = FO, colour = sampling_method)) +
  geom_jitter(alpha = 0.1, width = 0.3) +
  geom_point(data = sumdatSM, aes(y = meanFO), size = 2) +
  geom_errorbar(data = sumdatSM, aes(y = meanFO, ymin = meanFO - seFO, ymax = meanFO + seFO), width = 0.1) +
  xlab("Sampling method") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","grey")) +
   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))  


# Date
ggplot(seabird_data, aes(x = as.factor(date_median), y = FO)) +
  geom_point(data = sumdatDate, aes(y = meanFO), size = 2) +
  geom_errorbar(data = sumdatDate, aes(y = meanFO, ymin = meanFO - seFO, ymax = meanFO + seFO), width = 0.1) +
  xlab("Date of study (median)") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_discrete(breaks = c(1960, 1970, 1980, 1990, 2000, 2010, 2020))

# Diet
ggplot(seabird_data, aes(x = as.factor(main_diet), y = FO, colour = main_diet)) +
  geom_jitter(alpha = 0.1, width = 0.3) +
  geom_point(data = sumdatDiet, aes(y = meanFO), size = 2) +
  geom_errorbar(data = sumdatDiet, aes(y = meanFO, ymin = meanFO - seFO, ymax = meanFO + seFO), width = 0.1) +
  xlab("Main diet") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("#F94144","#F3722C",
                               "#F9C74F","#90BE6D",
                               "#577590")) +
  scale_x_discrete(labels = c("carrion", "cephalopods", "crustaceans", "fish", "other inverts")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

# Feeding method
ggplot(seabird_data, aes(x = as.factor(feeding_method), y = FO, colour = feeding_method)) +
  geom_jitter(alpha = 0.1, width = 0.3) +
  geom_point(data = sumdatFM, aes(y = meanFO), size = 2) +
  geom_errorbar(data = sumdatFM, aes(y = meanFO, ymin = meanFO - seFO, ymax = meanFO + seFO), width = 0.1) +
  xlab("Feeding method") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("#F94144","#F3722C","#F8961E",
                               "#F9C74F","#90BE6D","#43AA8B",
                               "#577590")) +
  scale_x_discrete(labels = c("aerial pursuit", "bottom feeding", "diving", 
                              "plunging", "scavenging", "skimming", "surface seizing")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))



```
