---
title: "Conclusions from a full model"
subtitle: "treating diet type as a random effect"
author: Andrew MacDonald with updates from Natalie Cooper
format: html
editor: source
---

We can learn a lot about this question just from looking carefully at one large model. 
In this model, diet is treated as a random effect.

```{r setup}
# Load libraries
library(brms)
library(tidyverse)
library(tidybayes)
library(ape)
library(here)

# Need to install CmdStan. Unpack and place folder into Phyloplastics folder
# https://github.com/stan-dev/cmdstan/releases/tag/v2.35.0
# Then in Terminal navigate to this folder and run "make build"
# For cmdstanr
# install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
library(cmdstanr)
# Provide path to where your CmdStan folder is
set_cmdstan_path(path = here("cmdstan-2.35.0/"))

# Add custom functions written by AM
source(here::here("analyses/functions.R"))
```

Read in data and tree, without NAs
```{r}
# Read in data
data_noNA <- readr::read_csv(file = here::here("data/plastic-data-aggregated-noNA.csv"))

# Read in tree
tree <- read.nexus(here::here("data/plastic-tree-noNA.nex"))

# Make this into a variance covariance matrix for the models
A.matrix <- ape::vcv.phylo(tree)

# Unite the diet categories
data_noNA_trait <- data_noNA |> 
  unite(col = "diet_type", Carrion.NonFishVertebrates:AS.Refuse)
```

Fit a full model

Note that we need to specify to use brms for bf as it is an internal function in geiger and can cause issues if this package is already loaded.
```{r model-fit, eval=TRUE}
model1_full_model_bf <- brms::bf(
  n_plastic | trials(n_total) ~ 1 + 
    mi(is_freeflying) + 
    Date_median + 
    (1 | Source) + 
    (1 | MarineRegion) + 
    (1 + Date_median| HBWv8.1_Binomial) + 
    (1               | gr(tips, cov = A)) + 
    (0 + Date_median | gr(tips, cov = A)) + 
    (1|diet_type)  + 
    (1 | Feeding.Method.Simple) +
    body_mass_2sd + 
    pelagic_specialist_CR,
  family = "binomial") + 
  brms::bf(is_freeflying | mi() ~ 1 + (1|Source)) + 
  set_rescor(FALSE)
```

Set priors

```{r}
model1_full_model_priors <- c(
  prior(normal(.5, .5), class = "Intercept", resp = "isfreeflying"),
  prior(exponential(3), class = "sd",        resp = "isfreeflying"),
  prior(exponential(1), class = "sd",        resp = "nplastic"),
  prior(normal(0, .2),  class = "Intercept", resp = "nplastic"),
  prior(normal(0, .2),  class = "b", coef = "Date_median", resp = "nplastic"),
  prior(normal(1, .5),  class = "b", coef = "miis_freeflying", resp = "nplastic"),
  prior(lkj(2),         class = "cor", group = "HBWv8.1_Binomial"),
  prior(normal(0, .2), class = "b", resp = "nplastic")
)
```

Run model
```{r}
# job::job({
model1_full_model <- brm(model1_full_model_bf, 
                          data = data_noNA_trait, 
                          data2 = list(A = A.matrix),
                          prior = model1_full_model_priors, 
                          backend = "cmdstanr", chains = 4, cores = 4,
                          file = here::here("output/model1_full_model.rds"),
                          file_refit = "never")
# })
```

```{r}
summary(model1_full_model)
```


## Exploratory figures

### Freeflying

```{r}
# full_two_phylo_params <- get_variables(full_two_phylo_brm)

# tail(full_two_phylo_param, 600)

freefly <- model1_full_model |> 
  spread_draws(b_isfreeflying_Intercept, r_Source__isfreeflying[Source,p])

freefly |> 
  mutate(fly = b_isfreeflying_Intercept  + r_Source__isfreeflying) |> 
  ungroup() |> 
  ggplot(aes(y = Source, x = fly)) + stat_pointinterval() + 
  geom_point(aes(y = Source, x = is_freeflying), 
             inherit.aes = FALSE,
             data = data_noNA_trait |> 
               select(Source, is_freeflying) |>
               distinct(), col = "red")

```


### Does plastic consumption change over time?

```{r}
conditional_effects(model1_full_model, "Date_median", resp = "nplastic")
```



Just the average effect of time:

```{r}

time_avg_bird <- tidyr::expand_grid(ROTL.tip.label = "Raphus cucullatus", 
                   TipLabel_HBWBLv6 = "Raphus cucullatus",
                   body_mass_2sd = 0,#c(-.5, 0, .5),
                   Date_median = seq(from = -40, to = 25, by = 5), 
                   source = "MacDonald 2024",
                   MarineRegion = "Atlantis", 
                   is_freeflying = 1,
                   n_total = 50,
                   pelagic_specialist_CR = 0) |> 
  tidybayes::add_epred_rvars(model1_full_model, 
                             resp = "nplastic", re_formula = NA,
                             allow_new_levels = FALSE
                             ) |> 
  mutate(FO = .epred / 50)

time_trend <- time_avg_bird |> 
  ggplot(aes(x = Date_median + 2000, ydist = FO)) + 
  stat_lineribbon() +
  scale_fill_brewer(palette = "Greens", direction = -1)  + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, 1)) + 
  labs(y = "Frequency of occurrance of plastic (FO)",
       x = "Year")

time_trend
```

Can I add points to this?

```{r}
obs_fo <- data_noNA_trait |> 
  group_by(Date_median, HBWv8.1_Binomial) |> 
  summarise(fo_sp_year = mean(n_plastic/n_total)) |> 
  ungroup() |> 
  mutate(Date_median = round(Date_median)) |> 
  group_by(Date_median) |> 
  summarize(avg = mean(fo_sp_year, na.rm = TRUE),
            sampsize = sum(!is.na(fo_sp_year)))

time_trend + 
  geom_point(aes(x = Date_median, y = avg, size = sampsize),
             inherit.aes = FALSE, data = obs_fo) #+ 

```

### Individual species curves




## Marine Region

```{r}
foo <- data_noNA_trait |> 
  modelr::data_grid(tips = "Raphus_cucullatus", 
                    HBWv8.1_Binomial = "Raphus cucullatus",
                    body_mass_2sd = 0,#c(-.5, 0, .5),
                    Date_median = 0, 
                    Source = "MacDonald 2024",
                    diet_type = "999",
                    MarineRegion = MarineRegion, 
                    is_freeflying = 1,
                    n_total = 50,
                    pelagic_specialist_CR = 0) |> 
  add_predicted_rvars(model1_full_model, 
                      re_formula = NULL, 
                      allow_new_levels = TRUE, 
                      sample_new_levels = "gaussian",
                      resp = "nplastic") 

foo |> 
  mutate(MarineRegion = forcats::fct_reorder(MarineRegion, .prediction)) |> 
  ggplot(aes(x = MarineRegion, ydist = .prediction/50)) + 
  stat_histinterval() + 
  theme_bw() + 
  labs(y = "Frequency of occurrance of plastic (FO)",
       x = "Marine region")

# ggsave("statistics_review/marine_region2.png")
  
```


```{r}
model1_full_model |> 
  tidybayes::spread_draws(r_MarineRegion__nplastic[region]) |> 
  tidybayes::compare_levels(r_MarineRegion__nplastic, by = region) |> 
  ungroup() |> 
  mutate(region = reorder(region, r_MarineRegion__nplastic)) |> 
  ggplot(aes(y = region, x = r_MarineRegion__nplastic)) + 
  ggdist::stat_halfeye()

ggsave(here::here("output/marine_region_compare.png"))
```


Southern Ocean and the North Pacific are two regions with more plastic consumption.


## Variance components

```{r}
brms::mcmc_plot(model1_full_model, variable = "^sd_*", regex = TRUE)
```

We have various random effects in this model.  We can calculate their relative contributions to unexplained variance. To do this, we have to square the standard deviations to get the variance of each random effect. Then we add all these variances together and divide by the sum. 

```{r}

relative_variance_table <- calculate_variance_comp(model1_full_model)

variance_component_fig <- relative_variance_table |> 
  ggplot(aes(y = varcomp, x = relvar)) + 
  stat_halfeye() + 
  labs(x = "Proportion of variance",
       y = "Variance component") + 
  theme_bw()

ggsave(filename = here::here("output/variance_components.png"),
       plot = variance_component_fig)

```

Here are the proportions as a table:

```{r}
relative_variance_table |> 
  group_by(varcomp) |> 
  summarize(mean = mean(relvar),
            sd = sd(relvar)) |> 
  arrange(desc(mean)) |> 
  knitr::kable()
```


## Time effect

Time curves across all species for the different regions

```{r}

# make a vector of all species traits, crossed with time

median_time_predictions <- data_noNA_trait |> 
  modelr::data_grid(nesting(TipLabel_HBWBLv6, 
                            ROTL.tip.label,
                            MarineRegion,
                            diet_type, 
                            Feeding.Method.Simple,
                            body_mass_2sd,
                            pelagic_specialist_CR),
                    is_freeflying = 1,
                    Source = "MacDonald2024",
                    Date_median = seq(from = -20, to = 20, length.out = 15),
                    n_total = 50
                    ) |> 
  add_epred_rvars(model1_full_model,
                  re_formula = ~ #(1 | Source) + 
                    (1 | MarineRegion) + 
                    (1 + Date_median | TipLabel_HBWBLv6) + 
                    (1 | gr(ROTL.tip.label, cov = A)) + 
                    (0 + Date_median | gr(ROTL.tip.label, cov = A)) + 
                    (1 | diet_type) + 
                    (1 | Feeding.Method.Simple), resp  = "nplastic") |> 
  mutate(med = median(.epred/50))

glimpse(median_time_predictions)

fo_time_predictions <- median_time_predictions |> 
  mutate(
    Feeding.Method.Simple = forcats::fct_recode(
      Feeding.Method.Simple,
      "Aerial pursuit" = "Aerial.Pursuit",
      "Bottom feeding" = "Bottom.Feeding",
      "Surface seizing" = "Surface.Seizing"),
    MarineRegion = str_replace(MarineRegion, " Ocean", "")) |> 
  ggplot(aes(x = Date_median + 2000, y = med, group = TipLabel_HBWBLv6)) + 
  geom_line(alpha = 0.7) + 
  facet_grid(MarineRegion ~ Feeding.Method.Simple) + 
  theme_bw() + 
  labs(x = "Year", y = "Predicted FO")

ggsave(filename = here::here("output/FO_time_predictions.png"), 
       plot = fo_time_predictions)
```



```{r}

## extract the slopes for both species names, merge


model1_full_model_param <- get_variables(model1_full_model)

# model1_full_model_param

average_slope <- tidybayes::spread_rvars(model1_full_model, b_nplastic_Date_median)

hbwb_effects <- model1_full_model |> 
  as_draws_rvars(variable = "^r_TipLabel_HBWBLv6__nplastic", regex = TRUE)

hbwb_rvars <- hbwb_effects |> 
  as.data.frame() |> 
  tibble::rownames_to_column(var = "TipLabel_HBWBLv6") |> 
  tibble::as_tibble()

rotl_rvars <-  model1_full_model |> 
  as_draws_rvars(variable = "^r_ROTL.tip.label__nplastic", regex = TRUE) |> 
  as.data.frame() |> 
  tibble::rownames_to_column(var = "ROTL.tip.label") |> 
  tibble::as_tibble()

# average slope

  
time_effect_sum <- data_noNA_trait |> 
  select(TipLabel_HBWBLv6, ROTL.tip.label) |> 
  distinct() |> 
  left_join(hbwb_rvars) |> 
  left_join(rotl_rvars) |> 
  mutate(sum_slope = r_TipLabel_HBWBLv6__nplastic.Date_median + r_ROTL.tip.label__nplastic.Date_median, 
         sum_inter =  r_TipLabel_HBWBLv6__nplastic.Intercept + r_ROTL.tip.label__nplastic.Intercept) |> 
  mutate(sum_slope = sum_slope + average_slope$b_nplastic_Date_median)

time_effect_sum |>
  ## add in the average
  mutate(TipLabel_HBWBLv6 = forcats::fct_reorder(TipLabel_HBWBLv6, sum_slope)) |> 
  ggplot(aes(xdist = sum_slope, y = TipLabel_HBWBLv6)) + 
  stat_pointinterval() + 
  geom_vline(xintercept = 0, col = "lightblue", size = 2) + 
  labs(y = "Species", x = "Effect of time (change in log odds per year)")
```

What proportion of species show evidence of an increase over time? 

```{r}
time_effect_sum |> 
  mutate(n_over_0 = sum(sum_slope > 0)/4000,
         mostly_direction = cut(n_over_0, breaks = c(0, 0.025, 0.975, 1),
                                labels = c("decline", "ns", "increase"))) |> 
  count(mostly_direction)
```


```{r}

## who are the most and least fastest changing
changing_sp <- time_effect_sum |> 
  select(TipLabel_HBWBLv6, sum_slope) |> 
  mutate(med = median(sum_slope)) |> 
  filter(med < -.05 | med > .05)

# pick these out of full dataset

obs_changing_sp <- data_noNA_trait |> 
  semi_join(changing_sp)

pred_changing_sp <- obs_changing_sp |> 
  modelr::data_grid(nesting(TipLabel_HBWBLv6, ROTL.tip.label, body_mass_2sd, pelagic_specialist_CR), 
                    Date_median = seq(from = -30, to = 25, length.out = 20), 
                    source = "MacDonald 2024",
                    MarineRegion = "Atlantis", 
                    is_freeflying = 1,
                    n_total = 50
                    ) |> 
  tidybayes::add_epred_rvars(model1_full_model, 
                             resp = "nplastic", 
                             re_formula = ~
                               (1 + Date_median| TipLabel_HBWBLv6) + 
                               (1               | gr(ROTL.tip.label, cov = A)) + 
                               (0 + Date_median | gr(ROTL.tip.label, cov = A)),
                             allow_new_levels = FALSE) |>
  mutate(FO = .epred/50)


pred_changing_sp |> 
  ggplot(aes(x = Date_median, ydist = FO)) + 
  stat_lineribbon() + 
  facet_wrap(~TipLabel_HBWBLv6) + 
  scale_fill_brewer(palette = "Greens") + 
  geom_point(aes(x = Date_median, y = n_plastic/n_total, size = n_total), 
             inherit.aes = FALSE, data = obs_changing_sp)

```

## fixed effects

```{r}
model1_full_model |> 
  fixef() |> 
  knitr::kable()
```

