---
title: "Exploring the data: distribution across taxonomy/phylogeny"
format: html
editor: visual
---

Load libraries

```{r}
library(ape)
library(tidyverse)
library(geiger)
library(tidytree)
library(treeio)
library(ggtree)
library(ggstance)
library(patchwork)
library(here)

# Helper functions for plotting
remove_y <- 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

remove_x <- 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# Colour palette
order_colours <- c("#ffd92f","#fc8d62","#8da0cb","#b3b3b3",
                "#a6d854","#66c2a5","#e5c494","#e78ac3")
```

Read in the tree and the data, including the full list of seabirds and whether they have been studied and if they have plastic if they have been studied.

```{r}
tree <- read.nexus(here("data/family-tree-2026.nex"))
full <- read_csv(here("data/full-species-list-2026.csv"))
taxa <- read_csv(here("data-raw/seabird-list-Clements2025.csv"))
```

Combine full data with taxonomic info and simplify the family names

```{r}
# Create a tips column in the taxa dataset
taxa <- taxa %>% mutate(tips = str_replace(Clements2025_Binomial, " ", "_"))
# join to plastic data
all <- left_join(full, taxa)

# Simplify family names
all <-
  all %>%
  mutate(Family = case_when(Clements2025_Family == "Anatidae (Ducks, Geese, and Waterfowl)" ~ "Anatidae",
                            Clements2025_Family == "Scolopacidae (Sandpipers and Allies)" ~ "Scolopacidae",    
                            Clements2025_Family == "Stercorariidae (Skuas and Jaegers)" ~ "Stercorariidae",      
                            Clements2025_Family == "Alcidae (Auks, Murres, and Puffins)" ~ "Alcidae",     
                            Clements2025_Family == "Laridae (Gulls, Terns, and Skimmers)" ~ "Laridae",    
                            Clements2025_Family == "Phaethontidae (Tropicbirds)" ~ "Phaethontidae",            
                            Clements2025_Family == "Gaviidae (Loons)" ~ "Gaviidae",                        
                            Clements2025_Family == "Spheniscidae (Penguins)" ~ "Spheniscidae",                
                            Clements2025_Family == "Diomedeidae (Albatrosses)" ~ "Diomedeidae",               
                            Clements2025_Family == "Oceanitidae (Southern Storm-Petrels)" ~ "Oceanitidae",    
                            Clements2025_Family == "Hydrobatidae (Northern Storm-Petrels)" ~ "Hydrobatidae",   
                            Clements2025_Family == "Procellariidae (Shearwaters and Petrels)"~ "Procellariidae",
                            Clements2025_Family == "Fregatidae (Frigatebirds)" ~ "Fregatidae",             
                            Clements2025_Family == "Sulidae (Boobies and Gannets)" ~ "Sulidae",         
                            Clements2025_Family == "Phalacrocoracidae (Cormorants and Shags)" ~ "Phalacrocoracidae", 
                            Clements2025_Family == "Pelecanidae (Pelicans)" ~ "Pelecanidae"))      

```

Now let's look at the numbers per family, and how the % coverage looks for each family relative to the studies and to the taxonomic distributions...

```{r}
# Get summary numbers per family
family_ds <-
  all %>%
  group_by(Clements2025_Order, Family) %>%
  # total studies per family
  summarise(total_studies = sum(n_studies),
           # total species per family
            total_taxa = n(),
  # total species with plastic per family
  total_plastic = length(which(plastic == "yes")),
  # total species studied per family
  total_species_studied = length(which(!is.na(plastic)))) %>%
  # percentage of taxa that have been studied
  mutate(percent_taxa = (total_species_studied/total_taxa) * 100) %>%
  mutate(percent_plastic = (total_plastic/total_species_studied) * 100) %>%
  rename(Order = Clements2025_Order)

# look at it 
family_ds

# Save results
#write_csv(family_ds, here("output/family-coverage.csv"))
```

Next let's do the same for the order, and how the % coverage looks for each family relative to the studies and to the taxonomic distributions...

```{r}
# Get summary numbers
order_ds <-
  all %>%
  group_by(Clements2025_Order) %>%
  # total studies per order
  summarise(total_studies = sum(n_studies),
           # total species per order
            total_taxa = n(),
  # total species with plastic per order
  total_plastic = length(which(plastic == "yes")),
  # total species studied per order
  total_species_studied = length(which(!is.na(plastic)))) %>%
  # percentage of taxa that have been studied
  mutate(percent_taxa = (total_species_studied/total_taxa) * 100) %>%
  mutate(percent_plastic = (total_plastic/total_species_studied) * 100) %>%
  rename(Order = Clements2025_Order)

# look at it 
order_ds

# Save results
#write_csv(order_ds, here("output/order-coverage.csv"))
```

### Plots of distributions

Family level plots

```{r}
# Number of studies
p1 <- 
  ggplot(family_ds, aes(y = total_studies, x = Family, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Family") +
  ylab("studies (n)") +
  theme(legend.position = "none")

# Number of species studied
p2 <-
  ggplot(family_ds, aes(y = total_taxa, x = Family, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Family") +
  ylab("species (n)") +
  theme(legend.position = "none")

# % coverage
p3 <- 
  ggplot(family_ds, aes(y = percent_taxa, x = Family, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Family") +
  ylab("% coverage") +
  theme(legend.position = "none") +
  remove_y

# % plastic
p4 <- 
  ggplot(family_ds, aes(y = percent_plastic, x = Family, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Family") +
  ylab("% with plastic") +
  theme(legend.position = "right") +
  remove_y

#p1 + p4
# Save plot
#ggsave(here("figures/family-level-plastic-coverage.png"), width = 10)

p2 + p3 + p4
# Save plot
#ggsave(here("figures/family-level-species-coverage.png"), width = 8)

```

Order level plots

```{r}
# Number of studies
p1 <- 
  ggplot(order_ds, aes(y = total_studies, x = Order, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Order") +
  ylab("studies (n)") +
  theme(legend.position = "none")

# Number of species studied
p2 <-
  ggplot(order_ds, aes(y = total_taxa, x = Order, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Order") +
  ylab("species (n)") +
  theme(legend.position = "none")

# % coverage
p3 <- 
  ggplot(order_ds, aes(y = percent_taxa, x = Order, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Order") +
  ylab("% coverage") +
  theme(legend.position = "none") +
  remove_y

# % plastic
p4 <- 
  ggplot(order_ds, aes(y = percent_plastic, x = Order, fill = Order)) + 
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = order_colours) +
  coord_flip() +
  theme_bw(base_size = 14) +
  #scale_y_continuous(breaks=seq(0,45,5)) +
  xlab("Order") +
  ylab("% with plastic") +
  theme(legend.position = "none") +
  remove_y

#p1 + p4
# Save plot
#ggsave(here("figures/order-level-plastic-coverage.png"), width = 10)

p2 + p3 + p4
# Save plot
#ggsave(here("figures/order-level-species-coverage.png"), height = 5)
```

### Plotting on the phylogeny
This creates Figure 1

Ensure the data is in the same order as the tree and make it into a dataframe

```{r}
# Reorder to match tree
family_ds <- family_ds[match(tree$tip.label, family_ds$Family),]

# Reorder the columns so Family is first
family_ds <-
  family_ds %>%
  select(Family, everything())

# Add rownames as family names
rownames(family_ds) <- family_ds$Family

# Make sure family_ds is a dataframe
family_ds <- as.data.frame(family_ds)
```

Make trees using ggtree.

Note that this code will throw lots of warnings about scales which can be ignored

```{r}
# Add colours for orders in the same order as the tree
tree_colours <- c("#ffd92f", rep("#fc8d62", 4), 
                  "#8da0cb","#b3b3b3", rep("#e78ac3",3),
                 "#e5c494", rep("#66c2a5", 4),"#a6d854")

# Make the tree base
base <- 
  ggtree(tree) +
  #geom_tiplab(geom = "text", offset = 5, colour = tree_colours)
  geom_tiplab(geom = "text", offset = 5)

# Plot species numbers as facet
p1 <- facet_plot(base, panel = 'Species (n)', 
                data = family_ds, 
                geom = geom_barh, 
                mapping = aes(x = total_taxa,
                              fill = Order), 
                stat = 'identity') +
  theme(legend.position = "none") +
  scale_fill_manual(values = order_colours)

# Plot % coverage as facet on top of old facet
p2 <- 
  facet_plot(p1, panel = 'Coverage (%)', 
                data = family_ds, 
                geom = geom_barh, 
                mapping = aes(x = percent_taxa,
                              fill = Order), 
                stat = 'identity') +
  theme_bw(base_size = 14) + 
  theme(legend.position = "right", legend.title = element_blank(),
        strip.background = element_rect(fill="white")) +
  #xlim_tree(200) + 
  #xlim_expand(c(0, 5), 'Coverage (%)') +
  remove_y +
  scale_fill_manual(values = order_colours)

# Plot % coverage as facet on top of old facet
p3 <- 
  facet_plot(p2, panel = 'Plastic (%)', 
                data = family_ds, 
                geom = geom_barh, 
                mapping = aes(x = percent_plastic,
                              fill = Order), 
                stat = 'identity') +
  theme_bw(base_size = 14) + 
  theme(legend.position = "right", legend.title = element_blank(),
        strip.background = element_rect(fill="white")) +
  xlim_tree(300) + 
  xlim_expand(c(0, 5), 'Plastic (%)') +
  remove_y +
  scale_fill_manual(values = order_colours)


p3

# Save the plot
#ggsave(here("figures/phylogeny-data-coverage.jpeg"), height = 7, width = 10)
```
