---
title: "MCMC glmm models"
format: html
editor: visual
---

```{r setup}
# Load libraries
library(MCMCglmm)
library(tidyverse)
library(tidybayes)
library(ape)
library(here)
library(coda)
```

Read in data and tree

```{r}
# Read in data
# Use read.csv because read_csv creates tibbles which MCMCglmm hates
plastic_data <- read.csv(file = here::here("data/plastic-data-aggregated_2025-09-24.csv"))

# Read in tree
plastic_tree <- read.nexus(here::here("data/seabird_tree_plastic_2025-09-15.nex"))
# Replace plotomies with 0 length branches
plastic_tree<- di2multi(plastic_tree)
# Make this into an inverse variance covariance matrix for the models
inv.phylo <- inverseA(plastic_tree, nodes = "TIPS", scale = TRUE)$Ainv

# Replace NAs for 0 for mixed freeflying
plastic_data <- plastic_data %>% mutate(freeflying = case_when(is.na(freeflying) ~ 0,
                                                               freeflying == 1 ~ 1))
# Remove family column as it causes MCMCglmm to fail
plastic_data <- plastic_data %>% select(-family)
```

### Fit a full model

We are modelling this as a binomial response, with number of individuals with plastics and total number of individuals jointly modelled. Need priors for each random effect.

```{r}
# Set up priors for MCMCglmm
# Inverse Wishart with V = 1 and nu = 0.02
# i.e. fairly uninformative priors
# G is the random effects
prior <- list(G = list(G1 = list(V = 1, nu = 0.02), # phylogeny
                       G2 = list(V = 1, nu = 0.02), # source
                       G3 = list(V = 1, nu = 0.02), # marine region
                       G4 = list(V = 1, nu = 0.02), # date
                       G5 = list(V = 1, nu = 0.02)), # freeflying
              R = list(V = 1, nu = 0.02)) # residuals

```

Set values for iterations, burnin and thinning interval
```{r}
nitt <- 10000000
thin <- 10000
burnin <- 100000
```

Fit the model

```{r}
model_full <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_agriculture + as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            date_median +
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)
```


Suggested priors?
#model1_full_model_priors <- c(
 # prior(normal(.5, .5), class = "Intercept", resp = "freeflying"),
 # prior(exponential(3), class = "sd",        resp = "freeflying"),
 # prior(exponential(1), class = "sd",        resp = "n_plastic"),
 # prior(normal(0, .2),  class = "Intercept", resp = "n_plastic"),
 # prior(normal(0, .2),  class = "b", coef = "Date_median", resp = "n_plastic"),
 # prior(normal(1, .5),  class = "b", coef = "mifreeflying", resp = "n_plastic"),
 # prior(lkj(2),         class = "cor", group = "Clements2024_Binomial"),
 # prior(normal(0, .2), class = "b", resp = "n_plastic")
)

Did the model converge? 

## Plot model diagnostics for MCMCglmm

For fixed effects

```{r}
plot(model_full$Sol)
```

For random effects

```{r}
plot(model_full$VCV)
```

```{r}
summary(model_full)
```

Calculate ESS

```{r}
effectiveSize(model_full$Sol[, 1:model_full$Fixed$nfl, drop = FALSE])[[1]] 
```

If this is above 200 so model has converged

Look for autocorrelation

```{r}
autocorr(model_full$VCV) # Autocorrelation close to 0
```

Look at summary of fixed effect results
```{r}
summary(model_full)$solutions
```

Look at G-structure and R-structure posterior modes

```{r}
posterior.mode(model_full$VCV)
```

Look at G-structure and R-structure 95% credible intervals

```{r}
HPDinterval(model_full$VCV)
```

DIC

```{r}
model_full$DIC 
```

How much variance is explained by phylogeny?

```{r}
summary(model_full)$Gcovariances # Look at phylogenetic variance component
```
```{r}
# Get lambda
lambda <- model_full$VCV[,'tips']/
  (model_full$VCV[,'tips'] + model_full$VCV[,'units'])

# Posterior mean of lambda
mean(lambda)
# Posterior mode of lambda
posterior.mode(lambda) 
# HPD interval 
HPDinterval(lambda)
```

# visualise
```{r}
library(bayesplot)

# Extract posterior samples
posterior_samples <- model_full$Sol

# Create a posterior plot
mcmc_areas(posterior_samples, prob = 0.8, prob_outer = 0.95) + labs(title = "Posterior Distributions", x = "Parameter Value")

```

# Prior testing

Set up new priors for MCMCglmm to ensure that the results are not due to prior specifications.

```{r}
priors_new <- list(G = list(G1 = list(V = 1, nu = 1), # phylogeny
                            G2 = list(V = 1, nu = 1), # source 
                            G3 = list(V = 1, nu = 1), # marine region
                            G4 = list(V = 1, nu = 1), # date
                            G5 = list(V = 1, nu = 1)), # freeflying
                            R = list(V = 1, nu = 1)) # residual variance


model_prior_test <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_agriculture + as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            date_median +
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = priors_new,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)                          
```

```{r}
priors_no_phylo <- list(G = list(G1 = list(V = 1, nu = 0.02), # phylogeny
                            #G2 = list(V = 1, nu = 1), # source 
                            G3 = list(V = 1, nu = 0.02), # marine region
                            G4 = list(V = 1, nu = 0.02), # date
                            G5 = list(V = 1, nu = 0.02)), # freeflying
                            R = list(V = 1, nu = 0.02)) # residual variance

model_no_phylo <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_agriculture + as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ source + 
                            marine_region + 
                            date_median +
                            freeflying,
                   #ginverse = list(tips = inv.phylo), 
                   prior = priors_no_phylo,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)  
```

Save models for later use
```{r}
saveRDS(model_full, "model_full.rds")
saveRDS(model_prior_test, "model_prior_test .rds")
saveRDS(model_no_phylo, "model_no_phylo.rds")
```
