---
title: "Summaries for results section"
format: html
editor: visual
---

## Summaries for results section

Load libraries

```{r}
library(tidyverse)
library(here)
```

Read in datasets

```{r}
# taxonomy info
species <- read_csv(here::here("data-raw/seabird-list-Clements2024_2025-09-10.csv"))
  
# plastic data post NA removal i.e. used in models
plastic <- read_csv(here::here("data/plastic-data-aggregated_2025-09-24.csv"))

# plastic data pre NA removal AND with non studied species added
all <- read_csv(here::here("data/full_species_list_2025-09-16.csv"))
```

How many papers in analyses?

```{r}
length(unique(plastic$source))
```

Manipulate to add taxonomy data

```{r}
# make a binomial column
all$clements2024_binomial <- gsub("_", " ", all$tips)

# combine with taxonomy info
all <- left_join(all, species, by = c("clements2024_binomial" = "Clements2024_Binomial"))
```

Summaries for data pre NA removal

```{r}
# total studies per family
plastic %>%
  group_by(order, family) %>%
  summarise(total_studies = n())
```

```{r}
# total studies per order
plastic %>%
  group_by(order) %>%
  summarise(total_studies = n())
```

```{r}
# total species per family
plastic %>%
  group_by(order, family) %>%
  summarise(total_species = length(unique(clements2024_binomial)))
```

```{r}
# total species per order
plastic %>%
  group_by(order) %>%
  summarise(total_species = length(unique(clements2024_binomial)))
```

```{r}
# total birds per family
plastic %>%
  group_by(order, family) %>%
  summarise(total_birds = sum(n_total))
```

```{r}
# total birds per order
plastic %>%
  group_by(order) %>%
  summarise(total_birds = sum(n_total))
```

```{r}
# total birds
plastic %>%
  summarise(total_birds = sum(n_total))
```

Summaries for data before we removed NAs

```{r}
# total studies per family
all %>%
  group_by(Clements2024_Order, Clements2024_Family) %>%
  summarise(total_studies = sum(n_studies))
```

```{r}
# total studies per order
all %>%
  group_by(Clements2024_Order) %>%
  summarise(total_studies = sum(n_studies))
```

```{r}
# total species per family
# this is a bit fiddly as we need to record species that do have data
all %>%
  group_by(Clements2024_Order, Clements2024_Family) %>%
  summarise(total_species = length(which(!is.na(total_n) == TRUE)))
```

```{r}
# total species per order
all %>%
  group_by(Clements2024_Order) %>%
  summarise(total_species = length(which(!is.na(total_n) == TRUE)))
```

```{r}
# total birds per family
all %>%
  group_by(Clements2024_Order, Clements2024_Family) %>%
  summarise(total_birds = sum(total_n, na.rm = TRUE))
```

```{r}
# total birds per order
all %>%
  group_by(Clements2024_Order) %>%
  summarise(total_birds = sum(total_n, na.rm = TRUE))
```

```{r}
# total birds
all %>%
  summarise(total_birds = sum(total_n, na.rm = TRUE))
```

How many birds have been studied?

```{r}
all %>% summarise(length(which(!is.na(total_n) == TRUE)))
```

How many birds have plastic?

```{r}
all %>% summarise(length(which(total_plastic > 0)))
```

Weighted mean FO for all seabirds

```{r}
all %>% 
  summarise(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE))
```

Weighted mean FO for orders

```{r}
all %>%
  group_by(Clements2024_Order) %>%
  summarise(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE))
```

Weighted mean FO for families

```{r}
all %>%
  group_by(Clements2024_Order, Clements2024_Family) %>%
  summarise(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE))
```

Weighted mean FO for species

```{r}
species_FO <-
  all %>%
  filter(!is.na(total_n)) %>%
  group_by(Clements2024_Order, Clements2024_Family, clements2024_binomial) %>%
  summarise(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE) *100,
            total_n) %>%
  mutate(Family = case_when(Clements2024_Family == "Anatidae (Ducks, Geese, and Waterfowl)" ~ "Anatidae",
                            Clements2024_Family == "Scolopacidae (Sandpipers and Allies)" ~ "Scolopacidae",    
                            Clements2024_Family == "Stercorariidae (Skuas and Jaegers)" ~ "Stercorariidae",      
                            Clements2024_Family == "Alcidae (Auks, Murres, and Puffins)" ~ "Alcidae",     
                            Clements2024_Family == "Laridae (Gulls, Terns, and Skimmers)" ~ "Laridae",    
                            Clements2024_Family == "Phaethontidae (Tropicbirds)" ~ "Phaethontidae",            
                            Clements2024_Family == "Gaviidae (Loons)" ~ "Gaviidae",                        
                            Clements2024_Family == "Spheniscidae (Penguins)" ~ "Spheniscidae",                
                            Clements2024_Family == "Diomedeidae (Albatrosses)" ~ "Diomedeidae",               
                            Clements2024_Family == "Oceanitidae (Southern Storm-Petrels)" ~ "Oceanitidae",    
                            Clements2024_Family == "Hydrobatidae (Northern Storm-Petrels)" ~ "Hydrobatidae",   
                            Clements2024_Family == "Procellariidae (Shearwaters and Petrels)"~ "Procellariidae",
                            Clements2024_Family == "Fregatidae (Frigatebirds)" ~ "Fregatidae",             
                            Clements2024_Family == "Sulidae (Boobies and Gannets)" ~ "Sulidae",         
                            Clements2024_Family == "Phalacrocoracidae (Cormorants and Shags)" ~ "Phalacrocoracidae", 
                            Clements2024_Family == "Pelecanidae (Pelicans)" ~ "Pelecanidae")) %>%
  ungroup() %>%
  arrange(-mean_species_FO_weighted) %>%
  select(Order = Clements2024_Order, Family, Species = clements2024_binomial, mean_species_FO_weighted)

#write_csv(species_FO, file = here::here("output/weighted-mean-species-FO.csv"))
```

```{r}
species_FO %>% filter(mean_species_FO_weighted >= 50) %>% summarise(n())
```

Which species are well studied?

We classify ‘well-studied’ species as those with ≥ 100 sampled birds across ≥ 3 records (i.e., multiple studies, years, locations and/or age classes).

```{r}
  all %>%
  filter(n_studies >=3 & total_n >= 100) %>% 
  summarise(n())
```

Which are well studied and susceptible?

```{r}
susceptible <-
  all %>%
  filter(n_studies >=3 & total_n >= 100) %>% 
  group_by(clements2024_binomial) %>%
  summarise(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE)) %>%
  filter(mean_species_FO_weighted > 0.50)

#write_csv(susceptible, file = here::here("output/susceptible.csv"))
```

Create dataset for plotting results figures

```{r}
results <- all %>%
  # add whether studied or not
  mutate(studied = 
           case_when(n_studies > 0  ~ "yes",                      
                     n_studies == 0 ~ "no")) %>%
  # add whether well studied or not
  mutate(well_studied = 
           case_when(n_studies >=3 & total_n >= 100 ~ "yes",                      
                     n_studies <3 | total_n < 100 ~ "no")) %>%
  # add weighted mean FO
  mutate(mean_species_FO_weighted = weighted.mean((total_plastic/total_n), total_n, na.rm = TRUE)) %>%
  mutate(Family = case_when(Clements2024_Family == "Anatidae (Ducks, Geese, and Waterfowl)" ~ "Anatidae",
                            Clements2024_Family == "Scolopacidae (Sandpipers and Allies)" ~ "Scolopacidae",    
                            Clements2024_Family == "Stercorariidae (Skuas and Jaegers)" ~ "Stercorariidae",      
                            Clements2024_Family == "Alcidae (Auks, Murres, and Puffins)" ~ "Alcidae",     
                            Clements2024_Family == "Laridae (Gulls, Terns, and Skimmers)" ~ "Laridae",    
                            Clements2024_Family == "Phaethontidae (Tropicbirds)" ~ "Phaethontidae",            
                            Clements2024_Family == "Gaviidae (Loons)" ~ "Gaviidae",                        
                            Clements2024_Family == "Spheniscidae (Penguins)" ~ "Spheniscidae",                
                            Clements2024_Family == "Diomedeidae (Albatrosses)" ~ "Diomedeidae",               
                            Clements2024_Family == "Oceanitidae (Southern Storm-Petrels)" ~ "Oceanitidae",    
                            Clements2024_Family == "Hydrobatidae (Northern Storm-Petrels)" ~ "Hydrobatidae",   
                            Clements2024_Family == "Procellariidae (Shearwaters and Petrels)"~ "Procellariidae",
                            Clements2024_Family == "Fregatidae (Frigatebirds)" ~ "Fregatidae",             
                            Clements2024_Family == "Sulidae (Boobies and Gannets)" ~ "Sulidae",         
                            Clements2024_Family == "Phalacrocoracidae (Cormorants and Shags)" ~ "Phalacrocoracidae", 
                            Clements2024_Family == "Pelecanidae (Pelicans)" ~ "Pelecanidae"))

  # replace NAs with 0s
  #tidyr::replace_na(list(0)
  #
#write_csv(results, file = here::here("output/results-for-tree-plotting.csv"))
```
