---
title: "MCMC glmm models"
format: html
editor: visual
---

Odds ratios?
poo <- summary(model_full)
exp(poo$solutions[,1])
An odds ratio greater than 1 indicates that the condition or event is more likely to occur in the first group. And an odds ratio less than 1 indicates that the condition or event is less likely to occur in the first group.


```{r setup}
# Load libraries
library(MCMCglmm)
library(tidyverse)
library(ape)
library(here)
library(coda)
library(geiger)

# set seed
set.seed(42)
```

Read in data and tree

```{r}
# Read in data
# Use read.csv because read_csv creates tibbles which MCMCglmm hates
plastic_data <- read.csv(file = here::here("data/plastic-data-aggregated_2026.csv"))

# Read in tree
plastic_tree <- read.nexus(here::here("data/seabird-tree-plastic-2026.nex"))
# Replace plotomies with 0 length branches
plastic_tree<- di2multi(plastic_tree)
# Make this into an inverse variance covariance matrix for the models
inv.phylo <- inverseA(plastic_tree, nodes = "TIPS", scale = TRUE)$Ainv

# Replace NAs for 0 for mixed freeflying
plastic_data <- plastic_data %>% mutate(freeflying = case_when(is.na(freeflying) ~ 0,
                                                               freeflying == 1 ~ 1))
# Rename family column as it causes MCMCglmm to fail
plastic_data <- plastic_data %>% rename(taxonomic_family = family)
```

### Scale and centre year

```{r}
plastic_data <- 
  plastic_data %>%
  mutate(year_centered = scale(date_median))
```

Look at multicollinearity...

```{r}
library(corrr)

# Remove variables not used in the analyses
plastic_data2 <- 
  plastic_data %>%
  dplyr::select(c(-body_mass_g, -anthropogenic_subsidy, -FO, -date_median, -as_agriculture))

# Correlation matrix
cor_matrix <- correlate(plastic_data2)
write_csv(cor_matrix, here("output/correlations.csv"))
```

### Fit a full model

We are modelling this as a binomial response, with number of individuals with plastics and total number of individuals jointly modelled. Need priors for each random effect.

```{r}
# Set up priors for MCMCglmm
# Inverse Wishart with V = 1 and nu = 0.02
# i.e. fairly uninformative priors
# G is the random effects
prior <- list(G = list(G1 = list(V = 1, nu = 0.02), # phylogeny
                       G2 = list(V = 1, nu = 0.02), # source
                       G3 = list(V = 1, nu = 0.02), # marine region
                       G4 = list(V = 1, nu = 0.02)), # freeflying
              R = list(V = 1, nu = 0.02)) # residuals

```

Set values for iterations, burnin and thinning interval

```{r}
nitt <- 100000
thin <- 100
burnin <- 5000
```

Fit the model. This takes a very long time to fit!

```{r}
model_full <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)
```

Did the model converge?

## Plot model diagnostics for MCMCglmm

For fixed effects

```{r}
plot(model_full$Sol)
```

For random effects

```{r}
plot(model_full$VCV)
```

```{r}
summary(model_full)
```

Calculate ESS

```{r}
effectiveSize(model_full$Sol[, 1:model_full$Fixed$nfl, drop = FALSE])[[1]] 
```

If this is above 200 so model has converged

Look for autocorrelation

```{r}
autocorr(model_full$VCV) # Autocorrelation close to 0
```

Look at summary of fixed effect results

```{r}
summary(model_full)$solutions
```

Look at G-structure and R-structure posterior modes

```{r}
posterior.mode(model_full$VCV)
```

Look at G-structure and R-structure 95% credible intervals

```{r}
HPDinterval(model_full$VCV)
```

DIC

```{r}
model_full$DIC 
```

How much variance is explained by random effects?

```{r}
summary(model_full)$Gcovariances
```

```{r}
# Get lambda
lambda <- model_full$VCV[,'tips']/
  (model_full$VCV[,'tips'] + model_full$VCV[,'units'])

# Posterior mean of lambda
mean(lambda)
# Posterior mode of lambda
posterior.mode(lambda) 
# Posterior median of lambda
median(lambda) 
# HPD interval 
HPDinterval(lambda)
```

# visualise

```{r}
library(bayesplot)

# Extract posterior samples
posterior_samples <- model_full$Sol

# Create a posterior plot
mcmc_intervals(posterior_samples, prob = 0.8, prob_outer = 0.95) + labs(title = "Posterior Distributions", x = "Parameter Value")

```

Save model for later use

```{r}
saveRDS(model_full, here::here("output/model-full_2026.rds"))
```

# Prior testing

Set up new priors for MCMCglmm to ensure that the results are not due to prior specifications.

```{r}
priors_new <- list(G = list(G1 = list(V = 1, nu = 1), # phylogeny
                            G2 = list(V = 1, nu = 1), # source 
                            G3 = list(V = 1, nu = 1), # marine region
                            G4 = list(V = 1, nu = 1)), # freeflying
                            R = list(V = 1, nu = 1)) # residual variance


model_prior_test <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = priors_new,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)                          
```

Save model for later use

```{r}
saveRDS(model_prior_test, here::here("output/model-prior-test_2026.rds"))
```

# Sensitivity analyses [added in response to reviewer comments]

## Remove pelagic specialists

```{r}
model_no_pelagic <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_no_pelagic, here::here("output/model-no-pelagic_2026.rds"))
```

```{r}
# Checks
plot(model_no_pelagic$Sol)
plot(model_no_pelagic$VCV)

# Outputs fixed effects
summary(model_no_pelagic)$solutions

# Random effects
summary(model_no_pelagic)$Gcovariances

# Get lambda
lambda_no_pelagic <- model_no_pelagic$VCV[,'tips']/
  (model_no_pelagic$VCV[,'tips'] + model_no_pelagic$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_no_pelagic)
# Posterior mode of lambda
posterior.mode(lambda_no_pelagic) 
# Posterior median of lambda
median(lambda_no_pelagic) 
# HPD interval 
HPDinterval(lambda_no_pelagic)
```

## Remove terrestrial feeding

```{r}
model_no_tf <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_no_tf, here::here("output/model-no-tf_2026.rds"))
```

```{r}
# Checks
plot(model_no_tf$Sol)
plot(model_no_tf$VCV)

# Outputs fixed effects
summary(model_no_tf)$solutions

# Random effects
summary(model_no_tf)$Gcovariances

# Get lambda
lambda_no_tf <- model_no_tf$VCV[,'tips']/
  (model_no_tf$VCV[,'tips'] + model_no_tf$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_no_tf)
# Posterior mode of lambda
posterior.mode(lambda_no_tf) 
# Posterior median of lambda
median(lambda_no_tf) 
# HPD interval 
HPDinterval(lambda_no_tf)
```

## Include nestbound birds

Prep the data
```{r}
# Read in data
# Use read.csv because read_csv creates tibbles which MCMCglmm hates
plastic_data_nb <- read.csv(file = here::here("data/plastic-data-aggregated-nest_2026.csv"))

# Read in tree
plastic_tree_nb <- read.nexus(here::here("data/seabird-tree-plastic-nest-2026.nex"))
# Replace plotomies with 0 length branches
plastic_tree_nb <- di2multi(plastic_tree_nb)
# Make this into an inverse variance covariance matrix for the models
inv.phylo_nb <- inverseA(plastic_tree_nb, nodes = "TIPS", scale = TRUE)$Ainv

# Replace NAs for 2 for mixed freeflying
plastic_data_nb <- plastic_data_nb %>% mutate(freeflying = case_when(is.na(freeflying) ~ 2,
                                                               freeflying == 1 ~ 1,
                                                               freeflying == 0 ~ 0))
# Remove family column as it causes MCMCglmm to fail
plastic_data_nb <- plastic_data_nb %>% select(-family)

plastic_data_nb <- 
  plastic_data_nb %>%
  mutate(year_centered = scale(date_median))
```

Run the model
```{r}
model_with_nb <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data_nb, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo_nb), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_no_tf, here::here("output/model-with-nb_2026.rds"))
```

```{r}
# Checks
plot(model_with_nb$Sol)
plot(model_with_nb$VCV)

# Outputs fixed effects
summary(model_with_nb)$solutions

# Random effects
summary(model_with_nb)$Gcovariances

# Get lambda
lambda_with_nb <- model_with_nb$VCV[,'tips']/
  (model_with_nb$VCV[,'tips'] + model_with_nb$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_with_nb)
# Posterior mode of lambda
posterior.mode(lambda_with_nb) 
# Posterior median of lambda
median(lambda_with_nb) 
# HPD interval 
HPDinterval(lambda_with_nb)
```

# Family and order level analyses

Only for families/orders with >30 studies that can be modelled, i.e.

ORDERS: Charadriiformes, Procellariiformes, Sphenisciformes and Suliformes

FAMILIES: Alcidae, Laridae, Diomedidae, Hydrobatidae, Oceanidae, Procellaridae, Spheniscidae (same as order Sphenisciformes), Phalacorcoridae, Sulidae.

## ORDERS

### Charadriiformes

Subset the data
```{r}
plastic_data_char <- 
  plastic_data %>%
  filter(order == "Charadriiformes")

# Subset the tree
remove <- name.check(data = plastic_data_char, phy = plastic_tree, 
           data.names = plastic_data_char$tips)

plastic_tree_char <- drop.tip(plastic_tree, remove$tree_not_data)

# Make this into an inverse variance covariance matrix for the models
inv.phylo_char <- inverseA(plastic_tree_char, nodes = "TIPS", scale = TRUE)$Ainv
```

Run the model

```{r}
model_char <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data_char, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo_char), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_char, here::here("output/model_Charadriiformes_2026.rds"))
```

```{r}
# Checks
plot(model_char$Sol)
plot(model_char$VCV)

# Outputs fixed effects
summary(model_char)$solutions

# Random effects
summary(model_char)$Gcovariances

# Get lambda
lambda_char <- model_char$VCV[,'tips']/
  (model_char$VCV[,'tips'] + model_char$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_char)
# Posterior mode of lambda
posterior.mode(lambda_char) 
# Posterior median of lambda
median(lambda_char) 
# HPD interval 
HPDinterval(lambda_char)
```

### Procellariiformes

Subset the data
```{r}
plastic_data_proc <- 
  plastic_data %>%
  filter(order == "Procellariiformes")

# Subset the tree
remove <- name.check(data = plastic_data_proc, phy = plastic_tree, 
           data.names = plastic_data_proc$tips)

plastic_tree_proc <- drop.tip(plastic_tree, remove$tree_not_data)

# Make this into an inverse variance covariance matrix for the models
inv.phylo_proc <- inverseA(plastic_tree_proc, nodes = "TIPS", scale = TRUE)$Ainv

```

Run the model

Exclude pelagic specialist because they all are
```{r}
model_proc <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + fish + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + bottom_feeding + diving + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data_proc, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo_proc), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_proc, here::here("output/model_Procellariiformes_2026.rds"))
```

```{r}
# Checks
plot(model_proc$Sol)
plot(model_proc$VCV)

# Outputs fixed effects
summary(model_proc)$solutions

# Random effects
summary(model_proc)$Gcovariances

# Get lambda
lambda_proc <- model_proc$VCV[,'tips']/
  (model_proc$VCV[,'tips'] + model_proc$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_proc)
# Posterior mode of lambda
posterior.mode(lambda_proc) 
# Posterior median of lambda
median(lambda_proc) 
# HPD interval 
HPDinterval(lambda_proc)
```

## FAMILIES

### Laridae

Subset the data
```{r}
plastic_data_lar <- 
  plastic_data %>%
  filter(taxonomic_family == "Laridae")

# Subset the tree
remove <- name.check(data = plastic_data_lar, phy = plastic_tree, 
           data.names = plastic_data_lar$tips)

plastic_tree_lar <- drop.tip(plastic_tree, remove$tree_not_data)

# Make this into an inverse variance covariance matrix for the models
inv.phylo_lar <- inverseA(plastic_tree_lar, nodes = "TIPS", scale = TRUE)$Ainv
```

Run the model
```{r}
model_lar <- MCMCglmm(cbind(n_plastic, n_total-n_plastic) ~ 
                         carrion + cephalopods + crustaceans + other_inverts + 
                         as_fishery + as_refuse +
                         + aerial_pursuit + plunging + 
                         scavenging + skimming + surface_seizing + terrestrial_feeding +
                         pelagic_specialist + 
                         body_mass_2sd +
                         year_centered -1, 
                   data = plastic_data_lar, 
                   family = "multinomial2",
                   random = ~ tips + 
                            source + 
                            marine_region + 
                            freeflying,
                   ginverse = list(tips = inv.phylo_lar), 
                   prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin,
                   verbose = FALSE)

# Save model
saveRDS(model_lar, here::here("output/model_Laridae_2026.rds"))
```

```{r}
# Checks
plot(model_lar$Sol)
plot(model_lar$VCV)

# Outputs fixed effects
summary(model_lar)$solutions

# Random effects
summary(model_lar)$Gcovariances

# Get lambda
lambda_lar <- model_lar$VCV[,'tips']/
  (model_lar$VCV[,'tips'] + model_lar$VCV[,'units'])

# Posterior mean of lambda
mean(lambda_lar)
# Posterior mode of lambda
posterior.mode(lambda_lar) 
# Posterior median of lambda
median(lambda_lar) 
# HPD interval 
HPDinterval(lambda_lar)
```
