---
title: "Plot data on phylogeny"
author: "Natalie Cooper"
---

### WEIGHTED MEAN FREQ of OCCURRENCE with bar plots around and coloured by order and/or suborder?
### groupOTU for colours

Load libraries
```{r}
library(ape) # for reading and manipulating trees
library(tidyverse) # for data manipulation
library(geiger) # for checking names in phylogeny and data match
library(ggtree) # for plotting trees
library(ggstance) # for geombarh function to make bars
library(ggnewscale) # for different scales in ggplot

# Helper functions for plotting
remove_y <- 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

Read in the phylogeny

```{r}
seabirdtree <- read.nexus(here("data/plastic-tree-full.nex"))
# look at it
str(seabirdtree)
```
Read in dataset
```{r}
plastic <- read_csv(here("data/plastic-data-full.csv"))

# Look at it
plastic
```

### Plotting on the phylogeny ###

Ensure the data is in the same order as the tree and make it into a dataframe
```{r}
# Reorder to match tree
plastic <- plastic[match(seabirdtree$tip.label, plastic$HBWv8.1_Binomial),]

# Create a new dataframe object that
# contains only the variable you want to plot
# with the species/tip names as rownames...
found <-
    plastic %>%
    mutate(plastic = as.factor(plastic)) %>%
    select(plastic) %>%
    as.data.frame()
  # Add rownames as species names  
  rownames(found) <- plastic$HBWv8.1_Binomial
  
number <-
    plastic %>%
    select(n) %>%
    as.data.frame()
  # Add rownames as species names  
  rownames(number) <- plastic$HBWv8.1_Binomial
```

Make trees using ggtree.

Note that this code will throw lots of warnings about scales which can be ignored
```{r}
# Make the tree base
seabirdtree$tip.label <- gsub("_", " ", seabirdtree$tip.label)
  
base <- 
  ggtree(seabirdtree, layout = "circular") +
  geom_tiplab(geom = "text", offset = 8, size = 1, fontface = "italic") +
  xlim(0,110)

# Plot stripes as facet
p1 <- gheatmap(base, found, offset = 0.5, width = 0.1,
           colnames = FALSE, color = "black", legend_title = "plastic",)+
# add fill colours using standard scale fill manual
  scale_fill_manual(values = c("cyan4", "gold"), na.value = "white") +
  labs(fill = "Plastic?")
p1

# Save plot
# ggsave(p1, file = here("phylogeny.pdf"), height = 10)
```

```{r}
# For adding more variables...

# If we do add number will need to cut out the one with 3500 birds as it skews the scale

# Now tell ggtree to add another plot but remove the existing scale
# Give this a name too - here I called in cool_plot2
p2 <- p1 + new_scale_fill()

# Now add the second heatmap
# Make sure the offset is bigger than the first one so it plots
# next to the first one - this might need a bit of fiddling to get right
p3 <- gheatmap(p2, number, offset = 3.5, width = 0.1,
         colnames = FALSE, color = "black") +
  scale_fill_viridis_c(option = "plasma")

p3
```

